---
title: "Tidy NEON Macroinvertebrates Data"
output: rmarkdown::html_vignette
author: Mariana Perez Rocha (with cosmetic edits by M.R. Helmus)
vignette: >
  %\VignetteIndexEntry{Tidy NEON Macroinvertebrates Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
#library(temporalNEON)
library(devtools)
library(neonUtilities) 
library(stringr)
library(dplyr)
library(lubridate)
library(here)
library(forcats)
devtools::load_all()
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Get Data
First we download and derive the data and then later tidy and explore the data.

### Macroinverts Data Product ID
```{r DPID}
my_dpid <- 'DP1.20120.001'
```

### Choose sites
All sites had to have > 2 years of sampling data as of 2019.


```{r site choice}

my_site_list <- c('ARIK','BARC','BLWA','CARI','COMO','CRAM','CUPE','GUIL','HOPB',
                  'KING','LECO','LEWI','MAYF','OKSR','POSE','PRIN','PRLA','PRPO',
                  'REDB','SUGG','TOMB','TOOK','WALK')

```

#### Note
This is how the reproducibility gets wonky. "used data from, `r length(my_site_list)` site."

### Download Data
The below code used to download the raw macroinverts files is not run during the package build. Don't ever eval it. 

Rerun this code to update the data.

The manuscript data are stored in \`data/macroinverts_raw.rda` and as a `csv` in  \`data-raw/submitted/`.

```{r download, eval=FALSE}
#download NEON data using API

# uncomment
#inv_allTabs <- loadByProduct(dpID = my_dpid, 
#                             site = my_site_list,
#                             package = "expanded", 
#                             nCores = 4, # likely not needed
#                             check.size = FALSE)

```

### Add taxonomy
Add to the downloaded list of data a table of taxonomy to clean up the data.
```{r taxon table}

# make ordered taxon_rank_list for a reference (subspecies is smallest rank, kingdom is largest)


taxon_rank_list_ordered <- c('kingdom', 
                             'subkingdom',
                             'infrakingdom', 
                             'superphylum', 
                             'phylum', 
                             'subphylum', 
                             'infraphylum',
                             'superdivision', 
                             'division', 
                             'subdivision', 
                             'infradivision', 
                             'parvdivision',
                             'superclass', 
                             'class',
                             'subclass',
                             'infraclass',
                             'superorder',
                             'order',
                             'suborder',
                             'infraorder',
                             'section',
                             'subsection',
                             'superfamily',
                             'family',
                             'subfamily',
                             'tribe',
                             'subtribe',
                             'genus',
                             'subgenus',
                             'speciesGroup',
                             'species',
                             'subspecies') %>% rev()
```

#### Note
This code is nor really used but could be useful. It gets a taxon table from the API and do save it in the main list.

```{r taxonomy download, eval=FALSE}
# full _taxon_table <- neonUtilities::getTaxonTable('MACROINVERTEBRATE')
```
### Load Data

0r

Save Data
```{r save data, eval=FALSE}
#---
# Uncomment to save a new version of the data

#inv_allTabs <- do.call(c, 
#                       list(inv_allTabs, # NEON download
#                            list(full_taxon_table = full_taxon_table), # NEON dl
#                            list(date_of_download = Sys.time()))) # unique id
#macroinverts_raw <- inv_allTabs
#use_data(macroinverts_raw, overwrite = TRUE)

#---

```

Load data and print it out.
```{r load data}
inv_allTabs <- macroinverts_raw
data.frame(tables = labels(inv_allTabs))
```

## Tidy Data
The raw data are stored in a list.

```{r check loaded list}
#names(inv_allTabs )
#names(inv_allTabs$inv_taxonomyProcessed)
# Check to see if all match
ifelse(length(setdiff(inv_allTabs$inv_taxonomyProcessed$sampleID, inv_allTabs$inv_fieldData$sampleID))>0,"STOP There is an error","All Good. Forward Ho!")
```

### Join Data
Join the cleaned taxonomy (processed taxonomy) with the field data (sampling data). 

Make a data set with a density variable `den`. This `den` variable is what is being tested for variance and stability. 
```{r join data, eval=TRUE}
#merge/join tables: processing macroinverts data to get to density/abundance 

# join cleaned taxonomy to sample data
inv_dat <- left_join(inv_allTabs$inv_taxonomyProcessed, 
                     inv_allTabs$inv_fieldData, 
                     by = c('sampleID')) %>% 
  mutate(den = estimatedTotalCount/benthicArea) %>% # make density
  mutate(scientificName = fct_explicit_na(scientificName)) %>% # explicit missing
  dplyr::filter(sampleCondition == "condition OK") # toss samples low quality
```
Tidy colnames.
```{r tidy inv_dat, eval=TRUE}
# remove duplicate col names and .x suffix
inv_dat <- inv_dat[,!grepl('\\.y',names(inv_dat))]
names(inv_dat) <- gsub('\\.x','',names(inv_dat))
head(inv_dat)
```
### Choose Taxonomic Scale
Toss all individuals not identified to the genus or lower taxonomic resolution (e.g., all individuals ided as Chironomidae are tossed).
```{r taxonomic filter}
# get genus and finer resolution using ordered taxon_rank_list
inv_dat$taxonRank_ordered <- factor(
  inv_dat$taxonRank,
  levels = taxon_rank_list_ordered,
  ordered = TRUE) 

# get all records that have rank <= genus, where genus is not NA or blank
inv_dat_fine <- inv_dat %>%
  filter(taxonRank_ordered <= 'genus') %>% # <= due to ordered factor
  filter(!is.na(genus), genus != '') # there are missing genera so toss them

# this table has all variables and it's not in a spread format yet
head(inv_dat_fine)


```

### Choose Spatial Scale

```{r spatial scale, eval=TRUE}

# grouping variables for aggregating density/abundance. Come back here if other vars are need in the final table.
#for now, it's easier to keep simple like this in order to get one entry per site per row (species per sites summed)
my_grouping_vars <- c('siteID','genus','collectDate')
```
### Choose Temporal Scale

```{r temporal scale}

# aggregate densities for each genus group, pull out year and month from collectDate, then
#excluding collectDate
inv_dat_aggregate_tidy <- inv_dat_fine %>%
  select(one_of(my_grouping_vars), den) %>%
  mutate(
    year = collectDate %>% lubridate::year(),
    month = collectDate %>% lubridate::month()
  ) %>%
  group_by_at(vars(my_grouping_vars, year, month)) %>%
  summarize(
    abundance = sum(den)) %>% 
  ungroup()

head(inv_dat_aggregate_tidy)

```
## Write Data
```{r}
## Codyn (year and month)
write.csv(inv_dat_aggregate , file = 
            'macroinverts_table_abundance_codyn.csv', row.names = F)

```




```{r sandbox DELETE, eval=FALSE}


###put in the formats required for codyn and BAT

inv_dat_aggregate <- inv_dat_fine %>%
  select(one_of(my_grouping_vars), den) %>%
  group_by_at(vars(my_grouping_vars)) %>%
  summarize(
    abundance = sum(den)) %>% 
  ungroup()

View(inv_dat_aggregate)


## Codyn (year and month)
write.csv(inv_dat_aggregate , file = 
            'macroinverts_table_abundance_codyn.csv', row.names = F)


##BAT year

agregate_year_BAT <- inv_dat_aggregate%>% 
  group_by(genus,year,siteID) %>%
  summarise(abund = mean(abundance))%>%
  spread(genus,abund, fill = 0)

View(agregate_year_BAT)

write.csv(agregate_year_BAT , file = 
            'macroinverts_table_abundance_year_BAT.csv', row.names = F)


## BAT month

# make wide first, filling with abundance, making the 'bout' (format year+month) to be used in BAT
inv_dat_wide <- inv_dat_aggregate %>%
  tidyr::spread(genus, abundance, fill = 0)%>% 
  unite(bout,'year':'month', na.rm = TRUE, remove = FALSE)%>%
  dplyr::select(-collectDate)


View(inv_dat_wide)

write.csv(inv_dat_wide, file = 
            'macroinverts_table_abundance_month_BAT.csv', row.names = F)
```
```{r UNK eval = FALSE}

# getting table of location into a data.frame (lat, long, elevation)
table_location <- inv_allTabs$inv_fieldData %>%
  select(namedLocation, decimalLatitude, decimalLongitude, elevation) %>%
  distinct() %>%
  rename(
    location_id = namedLocation,
    latitude = decimalLatitude,
    longitude = decimalLongitude
    ) 

write.csv(table_location, file = 
            'all_inverts_table_location.csv', row.names = F)


```

